---
title: "Dataset and patient comparison"
output: html_notebook
---

For each dataset, resp. patient, and cell type, we have genes that show 3'UTR shortening or lengthening. 

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
## setting working directory
opts_knit$set(root.dir = "~/Documents/gitlab/sc-polyasite")
```

```{r}
library(ggplot2)
library(dplyr)
library(reshape2)
library(UpSetR)
library(svglite)
library(biomaRt)
library(ggvenn)
```

Input arguments. Decide on one of three options.
```{r}
# Figure 4: cancer analysis
cell_types <- c('cancer_vs_alveolar_normal')
patients <- data.frame(patient = c('lambrechts', 'laughney'),
                       dataset = c('lambrechts','laughney'))
out_dir <- 'results/luad_matched/comparison/dataset_cancer'
idir <- 'auc_comparisons'
```

```{r}
# Figure 5: dataset analysis
cell_types <- c('alveolar', 'B_cell', 'endothelial', 'fibroblast', 'mast_cell', 'merged', 'myeloid', 'T_cell')
patients <- data.frame(patient = c('lambrechts', 'laughney'),
                       dataset = c('lambrechts','laughney'))
out_dir <- 'results/luad_matched/comparison/dataset_celltypes'
idir <- 'auc'
```

```{r}
# Figure 6: patient analysis
cell_types <- c('endothelial', 'fibroblast', 'merged', 'myeloid', 'T_cell')
patients <- data.frame(patient = c('patient3', 'patient4', 'patient6'),
                       dataset = c('lambrechts','lambrechts','lambrechts'))
out_dir <- 'results/luad_matched/comparison/dataset_patients'
idir <- 'auc'
```

General settings
```{r}
prefix <- 'cpm2'
```

# Load all TE ids  
```{r}
tes.shorter <- list()
tes.longer <- list()
for(i in 1:nrow(patients)){
  for(ct in cell_types){
    file <- file.path('results', 'luad_matched', patients[i, 'patient'], 
                      idir, 'analysis_out', paste0(prefix, '_TEs_shorter_IQR_', ct, '.tsv'))
    if(file.exists(file)){
      name <- paste(patients[i, 'patient'], ct, sep = ":")
      tes.shorter[[name]] <- read.table(file, sep = "\t", header = FALSE, stringsAsFactors = FALSE)[,1]
    }
    
    file <- file.path('results', 'luad_matched', patients[i, 'patient'], 
                      idir, 'analysis_out', paste0(prefix, '_TEs_longer_IQR_', ct, '.tsv'))
    if(file.exists(file)){
      name <- paste(patients[i, 'patient'], ct, sep = ":")
      tes.longer[[name]] <- read.table(file, sep = "\t", header = FALSE, stringsAsFactors = FALSE)[,1]
    }
  }
}
```

Obtain total number of genes.
Load set of TEs with sufficient expression.
```{r}
tes.filtered <- list()

for(i in 1:nrow(patients)){
  for(ct in cell_types){
    file <- file.path('results', 'luad_matched', patients[i, 'patient'], 
                      idir, 'analysis_out', paste0(prefix, '_TEs_filtered_', ct, '.tsv'))
    if(file.exists(file)){
      name <- paste(patients[i, 'patient'], ct, sep = ":")
      tes.filtered[[name]] <- read.table(file, sep = "\t", header = FALSE, stringsAsFactors = FALSE)[,1]
    }
  }
}
```

## Convert TE ids to gene ids
```{r}
new_config <- httr::config(ssl_verifypeer = FALSE)
httr::set_config(new_config, override = FALSE)

conv_ttog <- function(myvector, mytable){
  res <- sapply(myvector, function(x) mytable[mytable[,1] == x, 2][1])
  return(unname(res))
}

# select proper organism!
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
transcripts <- unique(unlist(c(lapply(tes.longer, sample), lapply(tes.shorter, sample))))
mouse_transcripts <- getBM(filters= "refseq_mrna", attributes= c("refseq_mrna","external_gene_name"),
               values = transcripts, mart= mart)

genes.shorter <- lapply(tes.shorter, function(x) conv_ttog(x, mouse_transcripts))
genes.longer <- lapply(tes.longer, function(x) conv_ttog(x, mouse_transcripts))

# omit NA and only unique genes
genes.shorter <- lapply(genes.shorter, function(x) unique(na.omit(x)))
genes.longer <- lapply(genes.longer, function(x) unique(na.omit(x)))
```

Replace TE ids by gene names in further analysis
```{r}
tes.shorter <- genes.shorter
tes.longer <- genes.longer
```

# Methods
```{r}
get_background <- function(ct, tes_list){
  tes.ct.background <- tes_list[grep(ct, names(tes_list))]
  tes.union <- Reduce(union, tes.ct.background)
  
  # convert to gene names
  transcripts <- getBM(filters= "refseq_mrna", attributes= c("refseq_mrna","external_gene_name"),
                 values = tes.union, mart= mart)
  genes.union <- unique(transcripts$external_gene_name)
  return(genes.union)
}

prob_overlap <- function(list1, list2, background){
  intersection <- intersect(list1, list2)
  x <- length(intersection)
  n <- length(unique(list1))
  K <- length(unique(list2))
  N <- length(background)
  odds_ratio <- x / ((n*K) / N)
  if(x < (n*K) / N){
    pval <- sum(dhyper(x = 0:x, m = K, n = N - K, k = n))
  } else {
    pval <- 1 - sum(dhyper(x = 0:(x-1), m = K, n = N - K, k = n))
  }
  if(pval < .Machine$double.eps){
    pval <- .Machine$double.eps
  }
  return(list(genes_intersection = intersection, odds_ratio = odds_ratio, pvalue = pval))
}
```

# General statistics

## Number of genes with 3'UTR shortening and lengthening per patient and cell type
```{r}
nr_tes_shorter <- sapply(tes.shorter, length)
nr_tes_longer <- sapply(tes.longer, length)
df <- data.frame(nr_tes_shorter, nr_tes_longer)
df$Patient <- sapply(rownames(df), function(x) strsplit(x, ":")[[1]][1])
df$cell_type <- sapply(rownames(df), function(x) strsplit(x, ":")[[1]][2])
# rename cell types
df$cell_type <- as.factor(df$cell_type)
levels(df$cell_type) <- sapply(levels(df$cell_type), function(x) gsub("_", " ", x))
# rename patient
df$Patient <- as.factor(df$Patient)

df.womerged <- filter(df, cell_type != 'merged')
```

Rename labels
```{r}
label_names <- c('shortening', 'lenthening')
names(label_names) <- colnames(df)[1:2]

label_names_dataset <- c('Lambrechts', 'Laughney')
names(label_names_dataset) <- levels(df$Patient)
```

Plot only merged.
Note: not applicable for cell_state_comparison!
```{r}
df.merged <- filter(df, cell_type == 'merged')
df.m.melt <- melt(df.merged, id.vars = c('Patient'), 
                  measure.vars = c('nr_tes_shorter', 'nr_tes_longer'))
# rename labels
levels(df.m.melt$variable) <- label_names
# NOTE: this only for dataset analysis!
levels(df.m.melt$Patient) <- label_names_dataset

# NOTE: change labs fill appropriately!
ggplot(df.m.melt, aes(x=variable, y=value, fill=Patient)) +
  geom_col(position = 'dodge') +
  theme_bw() +
  scale_fill_brewer(palette = "Dark2") +
  labs(y = '# genes', fill = "Dataset") +
  theme(axis.text.y = element_text(size=8),
        axis.text.x = element_text(size = 8, angle = 0, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=10))
ggsave(file.path(out_dir, 'bargraph_bulk_like_dodge.png'),
       device = "png", width = 8, height = 3.5, units = "cm")
```

Plot as bargraphs
```{r}
ggplot(df.womerged, aes(x=cell_type, y=nr_tes_shorter, fill=Patient)) + 
  geom_col(position = "dodge")  + 
  theme_bw() +
  labs(y = '# genes') + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text = element_text(size=8),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=10),
        legend.position = "none")
ggsave(file.path(out_dir, 'bargraph_genes_shorter.png'),
       device = "png", width = 7, height = 3.5, units = "cm")

ggplot(df.womerged, aes(x=cell_type, y=nr_tes_longer, fill=Patient)) + 
  geom_col(position = "dodge") +
  theme_bw() +
  labs(y = '# genes') + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.text = element_text(size=8),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=10),
        legend.position = "none")
ggsave(file.path(out_dir, 'bargraph_genes_longer.png'),
       device = "png", width = 7, height = 3.5, units = "cm")
```

Plot fraction of TEs shorter
```{r}
df.womerged <- mutate(df.womerged, fraction_shorter = nr_tes_shorter / (nr_tes_shorter + nr_tes_longer))

ggplot(df.womerged, aes(x=cell_type, y=fraction_shorter, color=Patient, group=Patient)) + 
  geom_line()  + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  geom_hline(yintercept = 0.5, color = "black") +
  lims(y=c(0,1)) +
  labs(y = 'fraction shortening') +
  theme(axis.text = element_text(size=8),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.title = element_blank(),
        legend.position = "none")
  
ggsave(file.path(out_dir, 'lineplot_fraction_genes_shorter.png'),
       device = "png", width = 7, height = 3.5, units = "cm")
```

## Overlaps

### Significance analysis

```{r}
ct <- cell_types[8]
```

Calculate significance
> for paired data only!

```{r}
# get cell type-specific background
genes.bg.ct <- get_background(ct, tes.filtered)
# Write statistical background out
write.table(genes.bg.ct,
              file = file.path(out_dir, paste0("bg_genes_", ct, ".tsv")), 
              sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

# For shortening events
tes.shorter.ct <- tes.shorter[grep(ct, names(tes.shorter))]
names(tes.shorter.ct) <- c('Lambrechts', 'Laughney')
# patient 3 and 6 only
# names(tes.shorter.ct) <- c('patient3', 'patient4', 'patient6')
# tes.shorter.ct[[2]] <- NULL

res.shorter.ct <- prob_overlap(tes.shorter.ct[[1]], tes.shorter.ct[[2]], genes.bg.ct)
# Write intersection genes out
write.table(res.shorter.ct$genes_intersection,
              file = file.path(out_dir, paste0("intersection_genes_shortening_", ct, ".tsv")), 
              sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

# For lengthening events
tes.longer.ct <- tes.longer[grep(ct, names(tes.longer))]
names(tes.longer.ct) <- c('Lambrechts', 'Laughney')
# patient 3 and 6 only
# names(tes.longer.ct) <- c('patient3', 'patient4', 'patient6')
# tes.longer.ct[[2]] <- NULL
res.longer.ct <- prob_overlap(tes.longer.ct[[1]], tes.longer.ct[[2]], genes.bg.ct)
# Write intersection genes out
write.table(res.longer.ct$genes_intersection,
              file = file.path(out_dir, paste0("intersection_genes_lengthening_", ct, ".tsv")), 
              sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
```
### Get overlap with ggVennDiagram
With ggvenn
```{r}

# colors if LB vs LN: c("#1B9E77", "#D95F02")
# if patient 3 vs 6: c("#1B9E77", "#7570B3")
cols <- c("#1B9E77", "#D95F02")
if(res.shorter.ct$pvalue <= .Machine$double.eps){
  text <- paste0("odds ratio = ", round(res.shorter.ct$odds_ratio, 2), 
                          "\np-value <= ", signif(res.shorter.ct$pvalue, 3))
} else {
  text <- paste0("odds ratio = ", round(res.shorter.ct$odds_ratio, 2), 
                          "\np-value = ", signif(res.shorter.ct$pvalue, 3))
}

p <- ggvenn(tes.shorter.ct, show_percentage = FALSE,
            fill_color = cols,
            text_size = 5,
            set_name_size = 5)+
  annotate("text", x = 0, y = -1.5, label = text, size = 5)
ggsave(file.path(out_dir, paste0('venn_genes_shorter_', ct, '.png')),
       plot = p, 
       device = "png", width = 7, height = 7, units = "cm", dpi = 300,
       scale = 1)

# lengthening
if(res.longer.ct$pvalue <= .Machine$double.eps){
  text <- paste0("odds ratio = ", round(res.longer.ct$odds_ratio, 2), 
                          "\np-value <= ", signif(res.longer.ct$pvalue, 3))
} else {
  text <- paste0("odds ratio = ", round(res.longer.ct$odds_ratio, 2), 
                          "\np-value = ", signif(res.longer.ct$pvalue, 3))  
}

p <- ggvenn(tes.longer.ct, show_percentage = FALSE,
            fill_color = cols,
            text_size = 5,
            set_name_size = 5)+
  annotate("text", x = 0, y = -1.5, label = text, size = 5)
ggsave(file.path(out_dir, paste0('venn_genes_longer_', ct, '.png')),
       plot = p, 
       device = "png", width = 7, height = 7, units = "cm", dpi = 300,
       scale = 1)
```
